---
title: "History8550 Digital Project Code All Iterations"
author: "Sally Mauldin"
date: "2023-11-14"
output:
  html_document: default
  pdf_document: default
---
---
<style type="text/css">
h6 { /* Header 6 */
 font-size: 10px;
 color: black;
 text-align: center
}
</style>

### Network of Colleges and Universities in South Carolina in the 19th Century

This visualization is a network of individuals associated with 22 higher education institutions in South Carolina in the 19th century. Each icon (node) represents one college/university, and the connections (edges) between icons indicate known associations, such as founders, presidents, and board members. 


```{r, message=FALSE, warning=FALSE, echo=FALSE}
#Load libraries
library(visNetwork)
library(igraph) 
library(dplyr)
library(tidyr)
library(ggplot2)
```

````{r, message=FALSE, warning=FALSE, echo=FALSE}
#Create edges and nodes from CSV files
edges <- read.csv("edges.version5.csv", stringsAsFactors = FALSE) 
edges <- as.data.frame(edges)

nodes <- read.csv("nodes.version6.csv", stringsAsFactors = FALSE) 
nodes <- as.data.frame(nodes)
````

````{r, message=FALSE, warning=FALSE, echo=FALSE}
#Create a column assigning colors to institutional relationships like President and Board Member
edges <- edges %>%
  mutate(title = edges$Institution.Roles[1:302]) %>%
  mutate(color = ifelse(edges$Institution.Roles == "Benefactor", "green",
                        ifelse(edges$Institution.Roles == "Board Member", "blue",
                               ifelse(edges$Institution.Roles == "President", "red",
                                      ifelse(edges$Institution.Roles == "Key Figure", "darkgreen",
                                                                                       ifelse(edges$Institution.Roles == "Founder", "purple",
                                                                                              ifelse(edges$Institution.Roles == "Namesake", "turquoise", "gray")))))))
````

````{r, message=FALSE, warning=FALSE, echo=FALSE}
#Customize the nodes by color in network. Nodes are all of the people and each of the 22 institutions. Total of 317 nodes.
nodes <- nodes %>%
  mutate(shape = ifelse(nodes$Political.Roles == "Legislator", "square",
                        ifelse(nodes$Political.Roles == "Founding Father,Legislator", "diamond",
                               ifelse(nodes$Political.Roles == "Legislator,Governor", "triangle",
                                      ifelse(nodes$Political.Roles == "Foreign Ambassador,Founding Father,Legislator", "diamond",
                                             ifelse(nodes$Political.Roles == "Lt. Governor,Legislator", "square",
                                                    ifelse(nodes$Political.Roles == "Founding Father,Governor,Legislator", "diamond",
                                                           ifelse(nodes$Political.Roles == "Governor,Legislator", "triangle",
                                                                  ifelse(nodes$Political.Roles == "Governor's Staff,Legislator", "square",
                                                                         ifelse(nodes$Political.Roles == "Governor", "star",
                                                                                ifelse(nodes$Political.Roles == "Founding Father,Judge,Governor,Legislator", "diamond",
                                                                                       ifelse(nodes$Political.Roles == "Foreign Ambassador,Legislator", "square",
                                                                                              ifelse(nodes$Political.Roles == "Legislator,Governor's Staff", "square",
                                                                                                     ifelse(nodes$Political.Roles == "Judge,Legislator", "square",
                                                                                                            ifelse(nodes$Political.Roles == "Foreign Ambassador", "open oval", "circle"
                                                                              )))))))))))))))%>%
  mutate(title = (nodes$id[1:317])) %>%
  mutate(color = ifelse(nodes$Type == "Person", "navy",
                        ifelse(nodes$Type == "Institution", "orange", "tan")))

```

```{r}
#Make data frames for edge and node legends
ledges <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 
```

```{r, echo=FALSE}
#Create network visualization
edges <- edges %>% rename(from = ID, to = ID.1)


visNetwork(nodes = nodes, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36) 
```


```{r}
#Filter by private institutions

nodes.private <- nodes %>%
  filter(Type.1 == "Private")

ledges.private <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.private <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 


visNetwork(nodes = nodes.private, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.private$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by public institutions
nodes.public <- nodes %>%
  filter(Type.1 == "Public" | Type.1 == "Proprietary")

ledges.public <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.public <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 


visNetwork(nodes = nodes.public, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.public$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by HBCU institutions

nodes.hbcu <- nodes %>%
  filter(HBCU == "TRUE")

ledges.HBCU <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.HBCU <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 

visNetwork(nodes = nodes.hbcu, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.hbcu$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by religious institutions

nodes.religious <- nodes %>%
  filter(Religious == "TRUE")

ledges.religious <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.religious <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 

visNetwork(nodes = nodes.religious, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.religious$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by female institutions

nodes.female <- nodes %>%
  filter(Female == "TRUE")

ledges.female <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.female <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 


visNetwork(nodes = nodes.female, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.female$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by military institutions

nodes.military <- nodes %>%
  filter(Military == "TRUE")

ledges.military <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.military <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 


visNetwork(nodes = nodes.military, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.military$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by integrated institutions

nodes.integrated <- nodes %>%
  filter(Integrated == "TRUE")

ledges.integrated <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.integrated <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 

visNetwork(nodes = nodes.integrated, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.integrated$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by co-educational institutions

nodes.coed <- nodes %>%
  filter(Co.Educational == "TRUE")

ledges.coed <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.coed <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 


visNetwork(nodes = nodes.coed, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.coed$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by land grant institutions

nodes.landgrant <- nodes %>%
  filter(Land_Grant == "TRUE")

ledges.landgrant <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.landgrant <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 



visNetwork(nodes = nodes.landgrant, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.landgrant$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by upstate region

nodes.upstate <- nodes %>%
  filter(Region == "Upstate")

ledges.upstate <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.upstate <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 


visNetwork(nodes = nodes.upstate, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.upstate$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by midlands region

nodes.midlands <- nodes %>%
  filter(Region == "Midlands")

ledges.midlands <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.midlands <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 



visNetwork(nodes = nodes.midlands, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.midlands$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by lowcountry region

nodes.lowcountry <- nodes %>%
  filter(Region == "Lowcountry")

ledges.lowcountry <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.lowcountry <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 


visNetwork(nodes = nodes.lowcountry, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.lowcountry$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by year founded - antebellum version

nodes.antebellumfounding <- nodes %>%
  filter(Year_Founded %in% (1770:1856))

ledges.antebellumfounding <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.antebellumfounding <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 

visNetwork(nodes = nodes.antebellumfounding, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.antebellumfounding$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```
```{r}
#Filter by year founded - Reconstruction and after

nodes.reconstructionfounding <- nodes %>%
  filter(Year_Founded %in% (1865:1900))

ledges.reconstructionfounding <- data.frame(color = c("green", "blue", "red", "darkgreen", "purple", "turquoise"),
                     label = c("Benefactor", "Board Member", "President", "Key Figure", "Founder", "Namesake"))

lnodes.reconstructionfounding <- data.frame(label = c("Person", "Institution"),
                    color = c("navy", "orange"),
                   ID = 1:2) 

visNetwork(nodes = nodes.reconstructionfounding, edges = edges, height = 650, width = 1100) %>%
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 12) %>% 
  visNodes(size = 36, 
           label = paste(nodes.reconstructionfounding$id),
           title = title,
           font = list("size" = 20)) %>%
  visEdges(color = list(color = edges$color, opacity = 0.2),
           hoverWidth = 20,
           selectionWidth = 36)
```


```{r}

```


```{r, echo=FALSE}


```
 
```{r}
#Create plot showing number of religious institutions by denomination

nodes.religiouschart <- nodes.religious %>%
filter(Type == "Institution") %>%
group_by(Denomination) %>%
summarize(count = n())

ggplot(nodes.religiouschart, aes(x = Denomination, y = count)) +
geom_col()

```

```{r}
#Create plot showing number of different types of political roles

nodes.political <- nodes %>%
filter(Type == "Person") %>%
  filter(Political.Roles != "NA") %>%
  group_by(Political.Roles) %>%
  summarize(count = n())

ggplot(nodes.political, aes(x = Political.Roles, y = count)) +
geom_col()
```

```{r}
#Create plot showing number of different types of political roles in private institutions

nodes.political.private <- nodes %>%
  filter(Type.1 == "Private") %>%
filter(Type == "Person") %>%
  filter(Political.Roles != "NA") %>%
  group_by(Political.Roles) %>%
  summarize(count = n())

ggplot(nodes.political.private, aes(x = Political.Roles, y = count)) +
geom_col()
```

```{r}
#Create plot showing number of different types of political roles in public institutions

nodes.political.public <- nodes %>%
  filter(Type.1 == "Public") %>%
filter(Type == "Person") %>%
  filter(Political.Roles != "NA") %>%
  group_by(Political.Roles) %>%
  summarize(count = n())

ggplot(nodes.political.public, aes(x = Political.Roles, y = count)) +
geom_col()
```


 
```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```


